---
title: "Statistical Analysis of Nodal Network Structures"
output:
  pdf_document: default
  html_document: default
date: "2025-03-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Dataset Information
Let's read in two datasets containing graph theory measures of patients who were sleeping while undergoing EEG. The target variable is Dreams, a binary (0,1) value that indicates if a patient reported a conscious experience. The data is messy, let's read it in, inspect it, and do some pre-processing. We will need a non-prametric test to compare the two groups, assumption violations are showcased on GitHub: .
```{r}
suppressPackageStartupMessages(library(readr))
df1 <- read_csv("C:/Users/User/Desktop/group1.csv", show_col_types = FALSE)
df2 <- read_csv("C:/Users/User/Desktop/group2.csv", show_col_types = FALSE)
#head(df1, 3);        
head(df2, 3)
```

```{r function}
process_combined_data <- function(df1, df2) {
  # Combine both data frames row-wise
  combined <- rbind(df1, df2)
  # Number of rows before dropping NA values
  n_before <- nrow(combined)
  # Remove rows with any NA values
  combined_clean <- tidyr::drop_na(combined)
  # Number of rows after dropping NA values
  n_after <- nrow(combined_clean)
  # Print the total number of rows dropped
  cat("Dropped", n_before - n_after, "rows containing NA values.\n")
  # Convert the Dreams column: 0 -> "No Dreams", 1 -> "Dreaming"
  combined_clean$Dreams <- factor(
    combined_clean$Dreams, 
    levels = c(0, 1), 
    labels = c("No Dreams", "Dreaming")
  )
  # Create a new column 'obs' to track observation number
  combined_clean$obs <- seq_len(nrow(combined_clean))
  # Return the cleaned, combined dataset
  return(combined_clean)
}
# Example usage:
processed_data <- process_combined_data(df1, df2)

```

```{r}
suppressPackageStartupMessages(library(ggstatsplot))
ggbetweenstats(
  data = processed_data,
  x = Dreams,             # 'Dreams' is a factor that indicates group membership
  y = degree,             # 'degree' is the graph measure to compare the two groups
  type = "nonparametric", # Nonparametric test is selected
  bf.message = FALSE,     # set to TRUE for postier probabilities
  xlab = "Patient Consciousness Status",
  ylab = "Nodal Degree",
  title = "Comparison of Degree by Dreams Group",
  messages = FALSE
)
```
Patients who reported dreaming had significantly lower nodal degree (a network connectivity measure) compared to those who did not, with a clinically meaningful effect, as evidenced by a large effect size (r = -0.57), non-overlapping confidence intervals (CI: [-0.66, -0.48]), and extreme statistical significance (p < 0.05)

